A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-01-20 paper/paper.tex
progress

We observe that the process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. Indeed, parametricity interprets a type $A$ by a relation $A_\kstar$ over this type, and a term $t: A$ as an evidence that $A_\kstar(t)(t)$ is inhabited. Technically, $\U$ is interpreted as the relation ${\U}_\kstar$, which takes two types $A_L$ and $A_R$ and returns the type $A_L \rightarrow A_R \rightarrow \U$, while a dependent function type $\Pi a: A.\, B$ is interpreted as the relation $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and returns the $\Pi$ type taking as arguments two copies of $a: A$ together with a line between the two copies. That is,

@@

In particular, for $X: \U$, applying the parametricity translation is about exhibiting a relation $X_\kstar$ which inhabits ${\U}_\kstar(X)(X)$ i.e. $X \rightarrow X \rightarrow \U$. In turn, the parametricity translation of $X_\kstar: X \rightarrow X \rightarrow \U$ is about finding an inhabitant $X_{\kstar\kstar}$ of $(X \rightarrow X \rightarrow \U)\_\kstar(X_\kstar)(X_\kstar)$ i.e. of:
%. Let us equivalently write the type of $X_1$ as $\Pi x_L: X_0.\, \Pi x_R: X_0.\, \U$. To define its translation, we need to triplicate each of $x_L: X_0$ and $x_R: X_0$ into two copies and a line connecting these copies, leading to the sequences of premises $x_{LL}: X_0$, $x_{LR}: X_0$, $x_{L\kstar}: X_1(x_{LL}$, $x_{LR})$ and $x_{RL}: X_0$, $x_{RL}: X_0$, $x_{R\kstar}: X_1(x_{RL})(x_{RR})$. Then, in the context of $x_L$ and $x_R$, interpreting $X_1(x_L)(x_R): \U$ itself leads to exhibiting an inhabitant of $X_1(x_{LL})(x_{LR}) \rightarrow X_1(x_{RL})(x_{RR}) \rightarrow \U$, therefore leading to exhibiting an inhabitant of:

\begin{align*}
  \Pi x_{LL}: X.\, \Pi x_{LR}: X\, X_\kstar(x_{LL})(x_{LR}) \\
  \rightarrow \Pi x_{RL}: X.\, \Pi x_{RR}: X.\, X_\kstar(x_{RL})(x_{RR}) \\ \rightarrow X_\kstar(x_{LL}, x_{RL}) \rightarrow X_\kstar(x_{LR})(x_{RR}) \\
  \rightarrow \U
\end{align*}

which hints us at how the sequence $X_0$, $X_1$, $X_2$ can be seen as a sequence of inhabitants of the iteration $X_0 \defeq X$, $X_1 \defeq X_\kstar$, $X_2 \defeq X_{\kstar\kstar}$ of parametricity applied to an initial $X: \U$. This tells us how the previous (informal) type given to $X_2$ could be rephrased so that it comes as the instance of a general recipe characterizing the type of all $X_i$.

>>>>

We observe that the process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. Indeed, parametricity interprets a type $A$ by a relation $A_\kstar$ over this type, and a term $t: A$ as an evidence that $A_\kstar(t)(t)$ is inhabited. Technically, $\U$ is interpreted as the relation ${\U}_\kstar$, which takes two types $A_L$ and $A_R$ and returns the type $A_L \rightarrow A_R \rightarrow \U$, while a dependent function type $\Pi a: A.\, B$ is interpreted as the relation $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and expresses that these functions map related arguments in $A$ to related arguments in $B$:

@@

In particular, for $X: \U$, applying the parametricity translation is about exhibiting a relation $X_\kstar$ which inhabits ${\U}_\kstar(X)(X)$ i.e. $X \rightarrow X \rightarrow \U$. In turn, the parametricity translation of $X_\kstar: X \rightarrow X \rightarrow \U$ is about finding an inhabitant $X_{\kstar\kstar}$ of $(X \rightarrow X \rightarrow \U)_\kstar(X_\kstar)(X_\kstar)$ i.e. of:

\begin{align*}
  \Pi x_{LL}: X.\, \Pi x_{LR}: X\, X_\kstar(x_{LL})(x_{LR}) \\
  \rightarrow \Pi x_{RL}: X.\, \Pi x_{RR}: X.\, X_\kstar(x_{RL})(x_{RR}) \\ \rightarrow X_\kstar(x_{LL}, x_{RL}) \rightarrow X_\kstar(x_{LR})(x_{RR})  \rightarrow \U
\end{align*}
which hints us at how the sequence $X_0$, $X_1$, $X_2$ can be seen as a sequence of inhabitants of the iteration

\begin{align*}
  X_0 &\defeq X:& \U \\
  X_1 &\defeq X_\kstar:& {\U}_\kstar(X)(X) \\
  X_2 &\defeq X_{\kstar\kstar}:& ({\U}_\kstar(X)(X))_{\kstar}(X_\kstar)(X_\kstar)
\end{align*}
of parametricity applied to an initial $X: \U$. This tells us how the previous (informal) type given to $X_2$ could be rephrased so that it comes as the instance of a general recipe characterizing the type of all $X_i$.
